---
output: github_document
---

<!-- badges: start -->
  [![R-CMD-check](https://github.com/skiptoniam/ppmData/workflows/R-CMD-check/badge.svg)](https://github.com/skiptoniam/ppmData/actions)
  [![Coverage Status](https://codecov.io/github/skiptoniam/ppmData/coverage.svg?branch=master)](https://codecov.io/github/skiptoniam/ppmData?branch=master)
<!-- badges: end -->
  
## ppmData is an R package that can be used to set up a quadrature scheme for spatial point processes modelling and could be used in other packages, like [ecomix](https://github.com/skiptoniam/ecomix) to run marked poisson point processes.

## Summary

The approach uses quasi-random sampling (Grafston & Tille, 2013, Foster et al., 2018) to generate a quadrature scheme for numerical approximation of a Poisson point process model (Berman & Turner 1992; Warton & Shepard 2010). Quasi-random sampling quadrature are a form of spatially-balanced survey design or point stratification that aims to reduce the frequency of placing samples close to each other (relative to pseudo-random or grid designs). A quasi-random quadrature design improves efficiency of background point sampling (and subsequent modelling) by reducing the amount of spatial auto-correlation between data implying that each sample is providing as much unique information as possible (Grafston & Tille, 2013, Foster et al., 2018) and thus reducing low errors for geostatistical prediction (Diggle & Ribeiro, 2007). Because the quasi-random design is not on a regular grid we use Dirichlet tessellation to generate polygons for each point in the quadrature scheme. Areal weights are then derived from these polygons. 

We note that this approach is different to a targeted background scheme for generating pseudo-absences. If the users intent is to reduce sighting biases via a targeted background scheme, we recommend that bias
is accounted for via covariates (distance to roads) or an offset (a known amount of effort) (e.g Warton et al., 2013; Renner et al, 2015).

## Using the ppmData package

### Installation
The `ppmData` package can be installed using `devtools` or `remotes` R packages.

```{r,eval=FALSE}
install.packages('devtools')
devtools::install_github('skiptoniam/ppmData')
```

### Example

Here is an example using a dataset from the `ppmData` package. We setup a quadrature scheme for the species *Tasmaphena sinclairi* located within Tasmania, Australia. 

```{r, message=FALSE}
library(ppmData)
path <- system.file("extdata", package = "ppmData")
lst <- list.files(path=path, pattern='*.tif',full.names = TRUE)
covariates <- rast(lst)
presences <- subset(snails,SpeciesID %in% "Tasmaphena sinclairi")
npoints <- 1000
ppmdata1 <- ppmData(npoints = npoints, presences=presences,
                    window = covariates[[1]], covariates=covariates)
```

Here we plot the quadrature scheme. The green points represent the known locations of *Tasmaphena sinclairi*. The black points represent the quadrature locations. Quasi-random quadrature where the integration points are generated using a quasi-random areal sample.

```{r fig1, fig.width=7, fig.height=4, fig.align='center', echo=TRUE}
plot(ppmdata1)
```
Once the `ppmData` object has been set up it is quiet easy to fit a point process
model in R. You can use something like `glm`, `gam` or even convert the 
`ppmData` object to work directly with `spatstat::ppm`. Here is a brief example 
of how you might fit a ppm using `glm`:

```{r, warning=FALSE}
ppp <- ppmdata1$ppmData
form  <- presence/weights ~ poly(annual_mean_precip, degree = 2) + 
                            poly(annual_mean_temp, degree = 2) + 
                            poly(distance_from_main_roads, degree = 2)

ft.ppm <- glm(formula = form, data = ppp,
              weights = as.numeric(ppp$weights),
              family = poisson())
```

One we have fitted the model, we might want to predict the relative intensity
across the window, this will give us something similar to the expected count of 
species presences per-cell.

```{r,fig.width=6, fig.height=6, fig.cap=""}
pred <- predict(covariates,
                ft.ppm,
                type="response")
plot(pred*prod(res(pred))) # scale response by area of cell.
points(presences[!duplicated(presences),1:2],pch=16,cex=0.75,col=gray(0.2,0.75))
```

### Quick comparison of quasirandom quadrature speeds for `ppmData` & `spatstat`

Using `ppmData` to generate a quasi-random scheme
```{r}
system.time(ppmdata2 <- ppmData(npoints = 100000,
                                presences=presences,
                                window = covariates[[1]],
                                covariates=covariates))
```

Using a simple spatstat example to generate a quasi-random scheme
```{r}
suppressPackageStartupMessages(library(spatstat))
data(simdat)
D <- rQuasi(100000,Window(simdat))
system.time(Q <- quadscheme(simdat,D,method="dirichlet", exact=FALSE))
```


## Code of Conduct
  
Please note that the ppmData project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

## References 

Diggle, P. J., P. J. Ribeiro, *Model-based Geostatistics*. Springer Series in Statistics. Springer, 2007.

Foster, S.D., Monk, J., Lawrence, E., Hayes, K.R., Hosack, G.R. and Przeslawski, R., 2018. *Statistical considerations for monitoring and sampling.* Field manuals for marine sampling to monitor Australian waters, pp.23-41.

Grafstrom, Anton, and Yves Tille. *Doubly balanced spatial sampling with spreading and restitution of auxiliary totals.* Environmetrics 24.2 (2013): 120-131.

Renner, I.W., Elith, J., Baddeley, A., Fithian, W., Hastie, T., Phillips, S.J., Popovic, G. and Warton, D.I., 2015. *Point process models for presence only analysis.* Methods in Ecology and Evolution, 6(4), p.366-379.

Warton, D. I., and L. C. Shepherd. *Poisson point process models solve the pseudo-absence problem for presence-only data in ecology.* The Annals of Applied Statistics 4.3 (2010): 1383-1402.

Warton, D.I., Renner, I.W. and Ramp, D., 2013. *Model-based control of observer bias for the analysis of presence-only data in ecology.* PloS one, 8(11), p.e79168.


